---
title: "4: Physical Properties of Lakes"
author: "Hydrologic Data Analysis | Kateri Salk"
date: "Fall 2019"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Lesson Objectives
1. Investigate the relationships between temperature, light, and oxygen in lakes by analyzing monitoring data
2. Analyze changes in lake surface temperatures over time
3. Apply data analytics skills to questions about physical properties of lakes
4. Communicate findings with peers through oral, visual, and written modes

## Opening Discussion
We analyzed temperature profiles and seasonal changes on Wednesday. What were some key takeaway points about thermal stratification and mixing?

There are two main, stratified layers in lakes. They mix within each other via traditional methods of mixing like convection, and mix between the two only when there is strong wind and a smaller density gradient between the layers.
Epi - top
Meta - middle
Hypo - bottom

Water is the most dense at 4 degrees Celsius

## Session Set Up
```{r, message = FALSE}
getwd()

library(tidyverse)
library(gridExtra)
library(cowplot)
library(lubridate)

NTLdata <- read.csv("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv")

theme_set(theme_classic())

```

## Data Wrangling and Exploration

### Investigate structure of dataset

```{r}
# Change sampledate to Date
NTLdata$sampledate <- as.Date(NTLdata$sampledate, "%m/%d/%y")

# Make three data frames, one for each lake
Pauldata <- filter(NTLdata, lakename == "Paul Lake")
Peterdata <- filter(NTLdata, lakename == "Peter Lake")
Tuesdaydata <- filter(NTLdata, lakename == "Tuesday Lake")

# Make a data frame to inspect 2016 data for Paul Lake
Pauldata.2016 <- filter(Pauldata, year4 == 2016)

```

### Data Visualization

Let's look at profiles of temperature, irradiance, and dissolved oxygen over the course of a season. We will use the `plot_grid` function from the cowplot package to put the graphs together on one plot. Pay attention to the differences among the formats of each plot and how these work together in the final plot. 

```{r}
Tempprofiles2016 <- 
  ggplot(Pauldata.2016, aes(x = temperature_C, y = depth, color = daynum)) +
  geom_point() +
  scale_y_reverse() +
  scale_x_continuous(position = "top") +
  scale_color_viridis_c(end = 0.8, option = "magma") +
  labs(x = expression("Temperature "(degree*C)), y = "Depth (m)") +
  theme(legend.position = "none")
print(Tempprofiles2016)

#You can specifiy a facet only if it's separate values in one column. R cannot facet across columns. You have to create as many graphs as there are columns, even if they will end up in the same graph.

Lightprofiles2016 <- 
  ggplot(Pauldata.2016, aes(x = irradianceWater, y = depth, color = daynum)) +
  geom_point() +
  scale_y_reverse() +
  scale_x_continuous(position = "top") +
  scale_color_viridis_c(end = 0.8, option = "magma") +
  labs(x = expression("Irradiance (W m"^-2*")"), y = "Depth (m)") +
  theme(legend.position = "none", 
        axis.text.y = element_blank(), axis.title.y = element_blank())
print(Lightprofiles2016)

Oxygenprofiles2016 <- 
  ggplot(Pauldata.2016, aes(x = dissolvedOxygen, y = depth, color = daynum)) +
  geom_point() +
  scale_y_reverse() +
  scale_x_continuous(position = "top") +
  scale_color_viridis_c(end = 0.8, option = "magma") +
  labs(x = expression("DO (mg L"^-1*")"), y = "Depth (m)", 
       color = "Julian Day") +
  theme(axis.text.y = element_blank(), axis.title.y = element_blank())
print(Oxygenprofiles2016)

Physicalprofiles2016 <- 
  plot_grid(Tempprofiles2016, Lightprofiles2016, Oxygenprofiles2016, 
            ncol = 3, rel_widths = c(1.25, 1, 1.5))
print(Physicalprofiles2016)

```

How do temperature profiles vary over the course of a season? What are the mechanisms for this pattern?

> Temperature generally rises over the course of the summer season - rising the most in the epilimnion and the least in the hypolimnion. The mechanisms for this pattern are the warming of the water by the sun/atmosphere.

How do irradiance profiles vary over the course of a season? What are the mechanisms for this pattern?

> Irradiance seems to peak mid-summer, which corresponds to the length of the day peaking around the same time. The deeper you go, the more light is scattered - this is exacerbated by the increase of density in water molecules.

How do dissolved oxygen profiles vary over the course of a season? What are the mechanisms for this pattern?

> Dissolved oxygen is at its highest at the beginning of the season and lowest at the end. It also seems to drop off considerably after a certain depth. 

> Colder water can hold more dissolved gases. 

> Bumps up in the metalimnion because phytoplankton love that area and photosynthesize, increasing the DO. [DEEP CHLOROPHYLL MAXIMUM] They hang out there because there is just enough sunlight to grow and there is still access to the nutrient-rich hypolimnion. 

> Bacterial respiration occurs in the hypolimnion and further depletes DO levels. 


## Data Analysis

### Long-term change in temperature
How much have surface temperatures increased over the period of study? Is this increase significant? Isolate surface depths and run the test for May, June, July, and August temperatures separately. Use a linear regression with year as the predictor variable and temperature as the response variable.

Steps: 

1. Add a column named "Month" to the data frame (hint: lubridate package)
2. Filter your data frame so that it only contains surface depths and months 5-8
3. Create 4 separate data frames, one for each month
4. Run a linear regression for each data frame (see instructions above)
5. For significant trends, calculate how many degrees the lake has warmed over the period of study
5. Create a ggplot of surface temps by date and facet by month. Add `geom_smooth(se = FALSE, method = lm)` to the plot to display the regression line.

```{r}
Pauldata$month <- month(as.Date(Pauldata$sampledate, origin = '1984-05-27'))
Pauldatafilter <- filter(Pauldata, depth == 0)
PauldatafilterMay <- filter(Pauldatafilter, month == 5)
PauldatafilterJune <- filter(Pauldatafilter, month ==6)
PauldatafilterJuly <- filter(Pauldatafilter, month == 7)
PauldatafilterAugust <- filter(Pauldatafilter, month == 8)

```

Summarize your results below. How would you explain your findings and the importance of these findings to a non-specialist audience while still employing the data and statistical results?

>

## Closing Discussion
How do warming surface waters impact the physical properties of a lake? How might this affect the ecology of the lake?

How do we reconcile large amounts of variability in long-term records?
